### 2019-07-08 2-Chapters/1-Chapter/chapter1.tex
Okay j'ai relu l'intro en français

% % Main problem
% \textbf{Main problems.}
%
The root of the problems motivating this thesis are the questions of global warming and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
In the last $150$ years, mankind has developed many different communication technologies, and since the late $1890$s, wireless communications between manufactured devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed worldwide, allowing a large range of different applications.
It is now a world-wide consensus that with the current trend of population increase and with the ongoing energy crisis, any newly deployed technology should be both cheap and energy efficient,
as well as adapted to serve a large number of people and devices.
%
Such IoT technologies should be able to adapt automatically to different environments and application contexts, and be as efficient as possible.
%
In addition to the usual Research and Development effort to design many different efficient radio access schemes, covering all the possible cases,
now the time has come to combine it with cheap and promising Machine Learning in order to (try to) attain the level of performance gain necessary for the IoT promises to become reality.

@@

This third kind of wireless systems are characterized by their decentralized nature,
where communications are initialized and regulated by the devices, not a centralized control systems
% due to the really limited information that the base station can send to the devices (in downlink), it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
Indeed, a central control requires signaling packets that have been identified as too heavy for these kinds of systems.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common problem is the strong constraint that such such IoT devices have on their power consumption, as most of them will be deployed without a direct power access and will run on a tiny battery, which lifespan should be maximized.
Most commercial IoT companies nowadays indeed sell a lifespan larger than $10$ years, like SIGFOX \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems (such as 4G/5G and WiFi).
%
Many different standards for IoT networks have been proposed in the recent years,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layer.
To quote some examples of IdO standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications (up-to \SI{2}{\meter}), while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications (up-to \SI{50}{\kilo\meter}).
We refer to the survey \cite{Centenaro16} for more details, and \cite{Azari18} or our recent article \cite{MoyBesson2019}, and references therein.

@@

\textbf{Spectrum Scarcity.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, and this limits the possibility of adding any new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below,
% only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz},
% while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz},
almost the entire spectrum
is allocated to various usages, that goes from maritime radionavigation (historically the first usage of radio telecommunication, \eg, in the Titanic), to space research, inter-satellite, mobile telephony and many other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the
\href{https://www.itu.int/en/Pages/default.aspx}{International Telecommunication Union} (see \href{https://www.itu.int/}{\texttt{www.ITU.int}}),
the \href{https://www.fcc.gov/}{Federal Communications Commission} in the United States of North America (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or the \href{https://cept.org}{European Conference of Postal and Telecommunications Administrations} in Europe (see \href{https://www.CEPT.org/}{\texttt{CEPT.org}}),
% or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}),
as well as different independent measurement campaigns, found that most radio frequency spectrum are inefficiently utilized.
This means that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{patil2011survey} for a survey on the worldwide spectrum utilization, and to \cite{valenta2010survey} for the situation in Europe.

@@

Cellular network bands are overloaded in most parts of the world, but other frequency bands (such as military, amateur radio and paging frequencies) are less utilized.
Independent studies performed in some countries confirmed that observation, and concluded that spectrum usage highly depends on both time and place, as shown by \cite{Lopez2009spectral} for instance.
Moreover, the fixed spectrum allocation prevents
the introduction of new services, especially for low-cost or small markets equipments.
% rarely used frequencies, like those assigned to specific services, from being used, even when any unlicensed users would not cause noticeable interference to the assigned service.
Therefore, in the last $15$ years, thanks to an active lobbying by the cognitive radio community,
regulatory bodies in the world have been considering whether to allow
a new wireless communication paradigm:
enable unlicensed users in licensed bands, if they would not cause any interference to the (paying) licensed users.
These initiatives are considered by the \textbf{Cognitive Radio},
field that we detail below, and especially for \textbf{Dynamic Spectrum Access (DSA)},
for which we refer to the surveys \cite{akyildiz2006next,garhwal2012survey} for more details.

@@

We illustrate below the idea of a \emph{learning cycle}, alternating between action and feedback,
in Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then by observing a reward $r(t)$, which is a certain measure of success of this action, produced by the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its rewards, by trials and errors (\ie, actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk or drive, to play a board or computer game, or discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

Focussing on one SU in an OSA network, it has to sequentially decide a channel to sense (in the set of channels, $[K]=\{1,\dots,K\}$), then it performs sensing, and finally it transmits in this channel if it was sensed free.
The goal of the SU is to minimize its energy consumption (we remind that we focus on ``green'' cognitive radio) and to maximize its uplink data rate, or equivalently, to maximize its number of successful transmission.
%
If the different channels are not uniformly used by the PU, and if we assume a stationary hypothesis on the PU traffic, then the goal of the SU boils down to exploring the different channels and exploiting the best ones.
This frames the Opportunistic Spectrum Access problem as an exploration/exploitation problem,
with a finite set of actions (the channels, also called \emph{arms}),
in a sequential action-then-feedback cycle (time steps are $t\in\N^*$),
under partial information feedback (\ie, the player receives feedback about only one channel over $K$).
These three hypotheses are the ones that restrict from the general ML framework to the specific MAB case (see Figure~\ref{fig:1:ReinforcementLearningCycle}).

\paragraph{MAB for OSA, and a short history of these researchs at the SCEE team.}
%
Previous works of our SCEE team showed that MAB can be used to model the OSA problem:
orthogonal frequency bands (or channels) are modeled by arms $k\in\{1,\dots,K\}$,
and the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$ is modeled by a reward of $r(t) \in \{0,1\}$.
Indeed, $r(t) = 1$ indicates that no Primary User was sensed (and thus a SU message can be sent), while a reward of $r(t)=0$ indicates that the channel $k$ is busy at during the time slot $t$, and no SU message should be sent, until the end of that slot.
%
This model was first studied by W. Jouini during his PhD thesis with C. Moy \cite{Jouini12PhD}, ten years ago, first in the article \cite{Jouini09} and later in \cite{Jouini10,Jouini12}.

Their works were among the first ones to propose to use Reinforcement Learning for Cognitive Radio and the OSA problem, especially the MAB model and the $\UCB_1$ algorithm,
along with the early works of Q. Zhao and her team, for instance with \cite{Liu08,Zhao10}.
%
Shortly after in $2014$, proof-of-concepts using real-world radio hardware and Software Defined Radio were developed by C. Moy and his student C. Robert \cite{RobertSDR2014,MoyWSR2014}, using USRP boards and the MATLAB/Simulink software.
In a second PhD thesis \cite{Modi17PhD}, N. Modi studied from $2014$ to $2017$ the impact of using MAB algorithms to optimize channel selections, on the battery life of a wireless device.
On the one hand, running a MAB algorithm such as UCB-like algorithms was proven to be useful and could bring significant improvement in terms of successful transmission rates, directly increasing the battery life of the device.
On the other hand, classical MAB algorithms tend to switch arms a lot of times, especially in the beginning of their learning process, and this induces a lot of dynamical hardware reconfigurations for the wireless device, as selecting a different channel requires a change in the radio hardware used by the device.
Each hardware reconfiguration costs energy for the device, and so quickly-switching algorithms lead to a reduction of the battery life.
The tradeoff between the two aspects is studied empirically in
\cite{modiDemo2016}.

@@

\paragraph{Limitations and specifities of IoT networks.}
%
The aforementioned literature have essentially shown that MAB algorithms can be applied with success for the OSA problem.
However, if we consider CR-ready devices that cannot perform sensing, such as low-cost and low energy-consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there are no longer a distinction between PU and SU.
In other words, there is no longer any priority between users in IoT networks.
%  (no more PU, all are SU).
%
The other specificities of IoT networks can be listed as follows,
and we refer to the survey \cite{Centenaro16} for more details.
\begin{itemize}\tightlist
    % \item
    % we consider here IoT networks that run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, run on low-quality hardware, and have limited software and power capabilities (\ie, tiny batteries).
    It means that even if they are not equipped with particular sensing capabilities,
    they are equipped with a transmitter as well as with a receiver (Tx and Rx, \ie, a transceiver),
    and they have low-complexity storage and computation capabilities (\ie, small embedded processors),
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low to very low duty cycles (few messages every minute to only a few every month).
\end{itemize}

@@

% \paragraph{Chapter~\ref{chapter:3}.}
%
Our exploration of the large number of MAB algorithms and models developed in the recent literature
has given us the ambition to write a single piece of software allowing anyone to easily implement new models and algorithms, in order to compare the existing ones and empirically explore the performances of newly proposed algorithms.
To tackle this goal, we developed a Python library of simulation of MAB problems.
%
To the best of our knowledge, we wrote the most exhaustive open-source library for bandits. Our library SMPyBandits is published online, under an open-source licence \cite{SMPyBanditsJMLR,SMPyBandits}, and hosted on \href{https://GitHub.com/SMPyBandits}{\texttt{GitHub.com/SMPyBandits}}.
We present in details its architecture and its features in Chapter~\ref{chapter:3}, along with different examples of its usage.
A full documentation is available online, as well as exhaustive instructions to reproduce the simulations discussed in the rest of this thesis.

@@

% \paragraph{Chapter~\ref{chapter:4}.}
%
We study two models of IoT networks in Chapter~\ref{chapter:4}.
The first model considers IoT devices that simply have data to regularly send to a base station (at unpredictable or random times).
Such devices use acknowledgements as feedback, in order to optimize their uplink communications, by accessing more the best channels (\ie, the channel less occupied by the surrounding traffic).
The goal of this first model is to increase the Quality of Service (QoS) of the application of such IoT network, by applying decentralized RL on the device side in order to reduce the rate of failed transmission.
The base station will thus receive more up-link packets from the devices being served, if they can successfully learn by themselves an efficient spectrum access scheme.
It implies that the network can accept more end-devices while maintaining the same QoS.
%
Our second model then considers packet retransmissions, and while the two concepts are similar, the application of an efficient decentralized learning algorithm now results in both an increased battery life for each of the devices, as well as an increased QoS for the entire network, as less retransmissions reduce the local spectrum load.

@@

On the one hand, in the OSA setting, we failed to obtain positive result for the \Selfish{} policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and analyzed by \cite{BoursierPerchet18} (Appendix~F).
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving significantly over previous state-of-the-art results.
Our algorithm also achieves a logarithmic number of collisions and arm switches, and allows a fixed group of $M$ devices to efficiently learn to use the $M$ best channels orthogonally for almost all their up-link communications.
%
This strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

    \item
    We also present the piece-wise stationary MAB model, in Chapter~\ref{chapter:6}, and a detailed literature review of the research on non-stationary MAB \cite{Besson2019GLRT,Besson2019Gretsi}.
    Following two recent works, we propose a new actively adaptive algorithm for the piece-wise stationary problem, \GLRklUCB, that achieves state-of-the-art performance.
    % - Literature review on non stationary models and algorithms, state-of-the-art for piece-wise stationary with our algorithm, GLR test + klUCB

@@

\item
    \emph{Multi-Armed Bandit Learning in IoT Networks and non-stationary settings},
    by Rémi Bonnefoi, \textbf{Lilian Besson}, Christophe Moy, Emilie Kaufmann \& Jacques Palicot.
    \emph{Conference on Cognitive Radio Oriented Wireless Networks},
    % \footnote{~CROWNCOM 2017, see \href{http://crowncom.org/2017}{\texttt{crowncom.org/2017}}},
    Lisboa, Portugal, Septembre $2017$,
    % \href{https://HAL.Inria.fr/hal-01575419}{\texttt{HAL.Inria.fr/hal-01575419}},
    \textbf{Best Paper Award}.
    See Section~\ref{sec:4:firstModel}.
    \cite{Bonnefoi17}

>>>>

% % Main problem
% \textbf{Main problems.}
%
The root of the problems motivating this thesis are the questions of global warming and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
In the last $150$ years, mankind has developed many different communication technologies, and since the late $1890$s, wireless communications between manufactured devices have been made possible, and more and more frequent in our lives.
With the advent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed worldwide, allowing a large range of different applications.
It is now a world-wide consensus that with the current trend of population increase and with the ongoing energy crisis, any newly deployed technology should be both cheap and energy efficient,
as well as adapted to serve a large number of people and devices.
%
Such IoT technologies should be able to adapt automatically to different environments and application contexts, and be as efficient as possible.
%
In addition to the usual Research and Development effort to design many different efficient radio access schemes, covering all the possible cases,
now the time has come to combine it with cheap and promising Machine Learning in order to (try to) attain the level of performance gain necessary for the IoT promises to become reality.

@@

This third kind of wireless systems are characterized by their decentralized nature,
where communications are initialized and regulated by the devices, not a centralized control systems
% due to the really limited information that the base station can send to the devices (in downlink), it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
Indeed, a central control requires signaling packets that have been identified as too heavy for these kinds of systems.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common problem is the strong constraint that such such IoT devices have on their power consumption, as most of them will be deployed without a direct power access and will run on a tiny battery, which lifespan should be maximized.
Most commercial IoT companies nowadays indeed sell a lifespan larger than $10$ years, like SIGFOX \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems (such as 4G/5G and WiFi).
%
Many different standards for IoT networks have been proposed in the recent years,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layer.
To quote some examples of IoT standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications (up-to \SI{2}{\meter}), while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications (up-to \SI{50}{\kilo\meter}).
We refer to the survey \cite{Centenaro16} for more details, and \cite{Azari18} or our recent article \cite{MoyBesson2019}, and references therein.

@@

\textbf{Spectrum Scarcity.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, and this limits the possibility of adding any new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below,
% only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz},
% while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz},
almost the entire spectrum
is allocated to various usages, that goes from maritime radio-navigation (historically the first usage of radio telecommunication, \eg, in the Titanic), to space research, inter-satellite, mobile telephony and many other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the
\href{https://www.itu.int/en/Pages/default.aspx}{International Telecommunication Union} (see \href{https://www.itu.int/}{\texttt{www.ITU.int}}),
the \href{https://www.fcc.gov/}{Federal Communications Commission} in the United States of North America (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or the \href{https://cept.org}{European Conference of Postal and Telecommunications Administrations} in Europe (see \href{https://www.CEPT.org/}{\texttt{CEPT.org}}),
% or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}),
as well as different independent measurement campaigns, found that most radio frequency spectrum are inefficiently utilized.
This means that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{patil2011survey} for a survey on the worldwide spectrum utilization, and to \cite{valenta2010survey} for the situation in Europe.

@@

Cellular network bands are overloaded in most parts of the world, but other frequency bands (such as military, amateur radio and paging frequencies) are less utilized.
Independent studies performed in some countries confirmed that observation, and concluded that spectrum usage highly depends on both time and place, as shown by \cite{Lopez2009spectral} for instance.
Moreover, the fixed spectrum allocation prevents
the introduction of new services, especially for low-cost or small markets equipment.
% rarely used frequencies, like those assigned to specific services, from being used, even when any unlicensed users would not cause noticeable interference to the assigned service.
Therefore, in the last $15$ years, thanks to an active lobbying by the cognitive radio community,
regulatory bodies in the world have been considering whether to allow
a new wireless communication paradigm:
enable unlicensed users in licensed bands, if they would not cause any interference to the (paying) licensed users.
These initiatives are considered by the \textbf{Cognitive Radio},
field that we detail below, and especially for \textbf{Dynamic Spectrum Access (DSA)},
for which we refer to the surveys \cite{akyildiz2006next,garhwal2012survey} for more details.

@@

We illustrate below the idea of a \emph{learning cycle}, alternating between action and feedback,
in Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then by observing a reward $r(t)$, which is a certain measure of success of this action, produced by the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its rewards, by trials and errors (\ie, actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk, to drive, to play a board or computer game, discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

Focusing on one SU in an OSA network, it has to sequentially decide a channel to sense (in the set of channels, $[K]=\{1,\dots,K\}$), then it performs sensing, and finally it transmits in this channel if it was sensed free.
The goal of the SU is to minimize its energy consumption (we remind that we focus on ``green'' cognitive radio) and to maximize its up-link data rate, or equivalently, to maximize its number of successful transmission.
%
If the different channels are not uniformly used by the PU, and if we assume a stationary hypothesis on the PU traffic, then the goal of the SU boils down to exploring the different channels and exploiting the best ones.
This frames the Opportunistic Spectrum Access problem as an exploration/exploitation problem,
with a finite set of actions (the channels, also called \emph{arms}),
in a sequential action-then-feedback cycle (time steps are $t\in\N^*$),
under partial information feedback (\ie, the player receives feedback about only one channel over $K$).
These three hypotheses are the ones that restrict from the general ML framework to the specific MAB case (see Figure~\ref{fig:1:ReinforcementLearningCycle}).

\paragraph{MAB for OSA, and a short history of these researches at the SCEE team.}
%
Previous works of our SCEE team showed that MAB can be used to model the OSA problem:
orthogonal frequency bands (or channels) are modeled by arms $k\in\{1,\dots,K\}$,
and the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$ is modeled by a reward of $r(t) \in \{0,1\}$.
Indeed, $r(t) = 1$ indicates that no Primary User was sensed (and thus a SU message can be sent), while a reward of $r(t)=0$ indicates that the channel $k$ is busy at during the time slot $t$, and no SU message should be sent, until the end of that slot.
%
This model was first studied by W. Jouini during his PhD thesis with C. Moy \cite{Jouini12PhD}, ten years ago, first in the article \cite{Jouini09} and later in \cite{Jouini10,Jouini12}.

Their works were among the first ones to propose to use Reinforcement Learning for Cognitive Radio and the OSA problem, especially the MAB model and the $\UCB_1$ algorithm,
along with the early works of Q. Zhao and her team, for instance with \cite{Liu08,Zhao10}.
%
Shortly after in $2014$, proof-of-concepts using real-world radio hardware and Software Defined Radio were developed by C. Moy and his student C. Robert \cite{RobertSDR2014,MoyWSR2014}, using USRP boards and the MATLAB/Simulink software.
In a second PhD thesis, N. Modi studied from $2014$ to $2017$ the impact on the battery life of a wireless device of using MAB algorithms to optimize channel selections \cite{Modi17PhD}.
On the one hand, running a MAB algorithm such as UCB-like algorithms was proven to be useful and could bring significant improvement in terms of successful transmission rates, directly increasing the battery life of the device.
On the other hand, classical MAB algorithms tend to switch arms a lot of times, especially in the beginning of their learning process, and this induces a lot of dynamical hardware reconfigurations for the wireless device, as selecting a different channel requires a change in the radio hardware used by the device.
Each hardware reconfiguration costs energy for the device, and so quickly-switching algorithms lead to a reduction of the battery life.
The trade-off between the two aspects is studied empirically in
\cite{modiDemo2016}.

@@

\paragraph{Limitations and specifities of IoT networks.}
%
The aforementioned literature have essentially shown that MAB algorithms can be applied with success for the OSA problem.
However, if we consider CR-ready devices that cannot perform sensing, such as low-cost and low energy-consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there are no longer a distinction between PU and SU.
In other words, there is no longer any priority between users in IoT networks.
%  (no more PU, all are SU).
%
The other specificities of IoT networks can be listed as follows,
and we refer to the survey \cite{Centenaro16} for more details.
\begin{itemize}\tightlist
    % \item
    % we consider here IoT networks that run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, run on low-quality hardware, and have limited software and power capabilities (\ie, tiny batteries).
    It means that even if they are not equipped with particular sensing capabilities,
    they are equipped with a transmitter as well as with a receiver (Tx and Rx, \ie, a transceiver),
    and they have low-complexity storage and computation capabilities (\ie, small embedded processors),
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low to very low duty cycles (few messages every minute to only a few every month), and are in charge of initiating their up-link communications.
\end{itemize}

@@

% \paragraph{Chapter~\ref{chapter:3}.}
%
Our exploration of the large number of MAB algorithms and models developed in the recent literature
has given us the ambition to write a single piece of software allowing anyone to easily implement new models and algorithms, in order to compare the existing ones and empirically explore the performances of newly proposed algorithms.
To tackle this goal, we developed a Python library of simulation of MAB problems.
%
To the best of our knowledge, we wrote the most exhaustive open-source library for bandits. Our library SMPyBandits is published online, under an open-source licence \cite{SMPyBanditsJMLR,SMPyBandits}, and hosted on \href{https://GitHub.com/SMPyBandits}{\texttt{GitHub.com/SMPyBandits}}.
We present in details its architecture and its features in Chapter~\ref{chapter:3}, along with different examples of its usage.
A full documentation is available online, as well as exhaustive instructions to reproduce the simulations presented in the rest of this thesis.

@@

% \paragraph{Chapter~\ref{chapter:4}.}
%
We study two models of IoT networks in Chapter~\ref{chapter:4}.
The first model considers IoT devices that simply have data to regularly send to a base station (at unpredictable or random times).
Such devices use acknowledgements as feedback, in order to optimize their up-link communications, by accessing more the best channels (\ie, the channel less occupied by the surrounding traffic).
The goal of this first model is to increase the Quality of Service (QoS) of the application of such IoT network, by applying decentralized RL on the device side in order to reduce the rate of failed transmission.
The base station will thus receive more up-link packets from the devices being served, if they can successfully learn by themselves an efficient spectrum access scheme.
It implies that the network can accept more end-devices while maintaining the same QoS.
%
Our second model then considers packet retransmissions, and while the two concepts are similar, the application of an efficient decentralized learning algorithm now results in both an increased battery life for each of the devices, as well as an increased QoS for the entire network, as less retransmissions reduce the local spectrum load.

@@

On the one hand, in the OSA setting, we failed to obtain positive result for the \Selfish{} policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and analyzed by other researchers in \cite{BoursierPerchet18} (Appendix~F).
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving significantly over previous state-of-the-art results.
Our algorithm also achieves a logarithmic number of collisions and arm switches, and allows a fixed group of $M$ devices to efficiently learn to use the $M$ best channels orthogonally for almost all their up-link communications.
%
This strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

    \item
    We also present the piece-wise stationary MAB model, in Chapter~\ref{chapter:6}, and a detailed literature review of the research on non-stationary MAB \cite{Besson2019GLRT,Besson2019Gretsi}.
    We propose a new actively adaptive algorithm for the piece-wise stationary problem, \GLRklUCB, that achieves state-of-the-art performance, while requiring no prior knowledge on the problem difficulty other than the number of break-points.
    % - Literature review on non stationary models and algorithms, state-of-the-art for piece-wise stationary with our algorithm, GLR test + klUCB

@@

\item
    \emph{Multi-Armed Bandit Learning in IoT Networks and non-stationary settings},
    by Rémi Bonnefoi, \textbf{Lilian Besson}, Christophe Moy, Emilie Kaufmann \& Jacques Palicot.
    \emph{Conference on Cognitive Radio Oriented Wireless Networks},
    % \footnote{~CROWNCOM 2017, see \href{http://crowncom.org/2017}{\texttt{crowncom.org/2017}}},
    Lisboa, Portugal, September $2017$,
    % \href{https://HAL.Inria.fr/hal-01575419}{\texttt{HAL.Inria.fr/hal-01575419}},
    \textbf{Best Paper Award}.
    See Section~\ref{sec:4:firstModel}.
    \cite{Bonnefoi17}

### 2019-07-08 2-Chapters/2-Chapter/chapter2.tex
Okay j'ai relu l'intro en français

This selection rule $A(t) \sim \cU(\arg\max\limits_{1\leq k \leq K} U_k^{\cA}(t))$ makes sense for the following reasons:
an arm $k$ is chosen at step $t$ because it has a large $\UCB_k(t)$, which can happen for two reasons.
$i)$ Because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high mean reward $\mu_k$,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
One can also observe that $\xi_k(t)\to\infty$ if $t\to\infty$ while $N_k(t)\ll\log(t)$, which can be used to prove that the $\UCB_1$ algorithm tries all arms at least $\Omega(\log(T))$ times (in average).
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{} defined in \eqref{eq:2:UCB_index} respectively represent \emph{exploitation} and \emph{exploration}, and summing them up is a natural way of dealing with the exploitation/exploration trade off.
The constant parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to choose $\alpha=1/2$ for optimal uniform performance (\ie, on all problems, \cite{Auer02}).

>>>>

This selection rule $A(t) \sim \cU(\arg\max\limits_{1\leq k \leq K} U_k^{\cA}(t))$ makes sense for the following reasons:
an arm $k$ is chosen at step $t$ because it has a large $\UCB_k(t)$, which can happen for two reasons.
$i)$ Because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high mean reward $\mu_k$,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
One can also observe that $\xi_k(t)\to\infty$ if $t\to\infty$ while $N_k(t)\ll\log(t)$, which can be used to prove that the $\UCB_1$ algorithm tries all arms at least $\Omega(\log(T))$ times (in average).
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{} defined in \eqref{eq:2:UCB_index} respectively represent \emph{exploitation} and \emph{exploration}, and summing them up is a natural way of dealing with the exploitation/exploration trade-off.
The constant parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to choose $\alpha=1/2$ for optimal uniform performance (\ie, on all problems, \cite{Auer02}).

### 2019-07-08 2-Chapters/3-Chapter/chapter3.tex
Okay j'ai relu l'intro en français

For the classical single-player stochastic MAB model, as defined in Chapter~\ref{chapter:2},
a stochastic MAB problem is defined by $K \geq 2$ distributions $(\nu_k)_{k\in[K]}$ (also called arms),
used to generate the \iid{} samples $Y_{k,t} \sim \nu_k, \forall t$.
%
An agent chooses arm $A(t)\in[K]$ at time $t$ and observes the reward $r(t) = Y_{A(t),t}$ without knowing the other (hidden) rewards (in opposition to the full-information setting where $(Y_{1,t},\dots,Y_{K,t})$ is observed).
%
Her goal is to maximize $\sum_{t=1}^T r(t)$, which require to trade off between exploring the unknown $K$ arms, exploiting the arm which is currently estimated to be the best one.

>>>>

For the classical single-player stochastic MAB model, as defined in Chapter~\ref{chapter:2},
a stochastic MAB problem is defined by $K \geq 2$ distributions $(\nu_k)_{k\in[K]}$ (also called arms),
used to generate the \iid{} samples $Y_{k,t} \sim \nu_k, \forall t$.
%
An agent chooses arm $A(t)\in[K]$ at time $t$ and observes the reward $r(t) = Y_{A(t),t}$ without knowing the other (hidden) rewards (in opposition to the full-information setting where $(Y_{1,t},\dots,Y_{K,t})$ is observed).
%
Her goal is to maximize $\sum_{t=1}^T r(t)$, which require to trade-off between exploring the unknown $K$ arms, exploiting the arm which is currently estimated to be the best one.
